---
title: "Airphoto Selection Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Airphoto Selection Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the full `fly` pipeline using bundled test data
from the Neexdzii Kwa (Morice River) floodplain near Houston, BC. The data
includes 1968 airphoto centroids at two scales (1:12,000 and 1:31,680).

```{r load}
library(fly)
library(sf)

centroids <- st_read(system.file("testdata/photo_centroids.gpkg", package = "fly"), quiet = TRUE)
aoi <- st_read(system.file("testdata/aoi.gpkg", package = "fly"), quiet = TRUE)
```

## Footprint estimation

`fly_footprint()` converts point centroids into rectangular polygons
representing estimated ground coverage. The standard 9" x 9" negative
produces a footprint width of `9 * scale * 0.0254` metres.

```{r footprint}
footprints <- fly_footprint(centroids)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40", main = "Photo footprints")
plot(st_geometry(footprints), border = "steelblue", add = TRUE)
plot(st_geometry(centroids), pch = 20, cex = 0.5, col = "red", add = TRUE)
```

## Spatial filtering

`fly_filter()` with `method = "footprint"` catches photos whose centroid
falls outside the AOI but whose footprint overlaps it — a common situation
with large-scale photos at the edge of the study area.

```{r filter}
fp_result <- fly_filter(centroids, aoi, method = "footprint")
ct_result <- fly_filter(centroids, aoi, method = "centroid")
cat("Footprint method:", nrow(fp_result), "photos\n")
cat("Centroid method: ", nrow(ct_result), "photos\n")
```

## Summary statistics

`fly_summary()` reports footprint dimensions and date ranges by scale.

```{r summary}
fly_summary(centroids)
```

## Coverage analysis

`fly_coverage()` computes what percentage of the AOI is covered by photo
footprints, grouped by any column.

```{r coverage}
fly_coverage(centroids, aoi, by = "scale")
```

## Greedy photo selection

`fly_select()` picks the minimum set of photos needed to reach a target
coverage, using a greedy set-cover algorithm that iteratively selects the
photo covering the most uncovered area.

```{r select}
selected <- fly_select(centroids, aoi, target_coverage = 0.80)
selected[, c("airp_id", "scale", "selection_order", "cumulative_coverage_pct")]
```

```{r plot-selected}
sel_fp <- fly_footprint(selected)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40",
     main = paste(nrow(selected), "photos selected"))
plot(st_geometry(sel_fp), border = "steelblue", col = adjustcolor("steelblue", 0.15), add = TRUE)
plot(st_geometry(selected), pch = 20, col = "red", add = TRUE)
```
