---
title: "Airphoto Selection Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Airphoto Selection Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the full `fly` pipeline using bundled test data
from the Upper Bulkley River floodplain near Houston, BC. The data
includes 1968 airphoto centroids at two scales (1:12,000 and 1:31,680).

```{r load}
library(fly)
library(sf)

centroids <- st_read(system.file("testdata/photo_centroids.gpkg", package = "fly"), quiet = TRUE)
aoi <- st_read(system.file("testdata/aoi.gpkg", package = "fly"), quiet = TRUE)
```

## Footprint estimation

`fly_footprint()` converts point centroids into rectangular polygons
representing estimated ground coverage. The standard 9" x 9" negative
produces a footprint width of `9 * scale * 0.0254` metres.

```{r footprint}
footprints <- fly_footprint(centroids)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40", main = "Photo footprints")
plot(st_geometry(footprints), border = "steelblue", add = TRUE)
plot(st_geometry(centroids), pch = 20, cex = 0.5, col = "red", add = TRUE)
```

## Spatial filtering

`fly_filter()` with `method = "footprint"` catches photos whose centroid
falls outside the AOI but whose footprint overlaps it — a common situation
with large-scale photos at the edge of the study area.

```{r filter}
fp_result <- fly_filter(centroids, aoi, method = "footprint")
ct_result <- fly_filter(centroids, aoi, method = "centroid")
cat("Footprint method:", nrow(fp_result), "photos\n")
cat("Centroid method: ", nrow(ct_result), "photos\n")
```

## Summary statistics

`fly_summary()` reports footprint dimensions and date ranges by scale.

```{r summary}
fly_summary(centroids)
```

## Coverage analysis

`fly_coverage()` computes what percentage of the AOI is covered by photo
footprints, grouped by any column.

```{r coverage}
fly_coverage(centroids, aoi, by = "scale")
```

## Greedy photo selection

`fly_select()` picks the minimum set of photos needed to reach a target
coverage, using a greedy set-cover algorithm that iteratively selects the
photo covering the most uncovered area.

```{r select}
selected <- fly_select(centroids, aoi, target_coverage = 0.80)
selected[, c("airp_id", "scale", "selection_order", "cumulative_coverage_pct")]
```

```{r plot-selected}
sel_fp <- fly_footprint(selected)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40",
     main = paste(nrow(selected), "photos selected (greedy)"))
plot(st_geometry(sel_fp), border = "steelblue", col = adjustcolor("steelblue", 0.15), add = TRUE)
plot(st_geometry(selected), pch = 20, col = "red", add = TRUE)
```

## Priority selection: best resolution first

In practice you often want **all** photos at the best available resolution,
then backfill uncovered area with coarser scales. This combines `fly_select()`
with a priority loop over scales.

```{r priority}
sf_use_s2(FALSE)
scales_priority <- c("1:12000", "1:31680")
target_coverage <- 0.80

aoi_albers <- st_transform(aoi, 3005) |> st_union() |> st_make_valid()
aoi_area <- as.numeric(st_area(aoi_albers))
selected_all <- NULL
remaining_aoi <- aoi_albers

for (i in seq_along(scales_priority)) {
  sc <- scales_priority[i]
  photos_sc <- centroids[centroids$scale == sc, ]
  if (nrow(photos_sc) == 0) next

  if (i == 1) {
    # Best resolution: take ALL photos
    cat(sc, ": taking all", nrow(photos_sc), "photos\n")
    sel <- photos_sc
    sel$selection_order <- seq_len(nrow(sel))
    sel$cumulative_coverage_pct <- NA_real_
  } else {
    # Coarser scales: greedy select on remaining uncovered area
    remaining_sf <- st_sf(geometry = st_geometry(remaining_aoi)) |>
      st_transform(4326) |> st_make_valid()
    sel <- fly_select(photos_sc, remaining_sf, target_coverage = target_coverage)
  }

  if (nrow(sel) > 0) {
    fp <- fly_footprint(sel) |> st_transform(3005)
    fp_union <- st_union(fp) |> st_make_valid()
    remaining_aoi <- tryCatch(
      st_difference(remaining_aoi, fp_union) |> st_make_valid(),
      error = function(e) remaining_aoi
    )
    covered_pct <- 1 - as.numeric(st_area(remaining_aoi)) / aoi_area
    cat("  -> ", nrow(sel), " photos at ", sc,
        " (cumulative: ", round(covered_pct * 100, 1), "%)\n")
    sel$priority_scale <- sc
    selected_all <- rbind(selected_all, sel)
  }
}

cat("\nTotal:", nrow(selected_all), "photos\n")
as.data.frame(table(selected_all$priority_scale))
```

```{r plot-priority}
sel_fp <- fly_footprint(selected_all)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40",
     main = paste(nrow(selected_all), "photos (priority selection)"))
# Color by scale
cols <- ifelse(selected_all$priority_scale == "1:12000", "steelblue", "darkorange")
border_cols <- cols
fill_cols <- adjustcolor(cols, 0.15)
for (j in seq_len(nrow(sel_fp))) {
  plot(st_geometry(sel_fp[j, ]), border = border_cols[j],
       col = fill_cols[j], add = TRUE)
}
legend("topright", legend = scales_priority,
       fill = adjustcolor(c("steelblue", "darkorange"), 0.3),
       border = c("steelblue", "darkorange"), bty = "n")
```
