---
title: "Airphoto Selection Pipeline"
output:
  bookdown::html_document2:
    number_sections: true
    base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Airphoto Selection Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the full `fly` pipeline using bundled test data
from the Upper Bulkley River floodplain near Houston, BC. The data
includes 1968 airphoto centroids at two scales (1:12,000 and 1:31,680).

```{r load}
library(fly)
library(sf)

centroids <- st_read(system.file("testdata/photo_centroids.gpkg", package = "fly"), quiet = TRUE)
aoi <- st_read(system.file("testdata/aoi.gpkg", package = "fly"), quiet = TRUE)
```

# Footprint estimation

`fly_footprint()` converts point centroids into rectangular polygons
representing estimated ground coverage. The standard 9" x 9" (228 mm) negative
used by BC aerial survey cameras produces a footprint width of
`9 * scale * 0.0254` metres, which is the default (`negative_size = 9`).

Ground width scales with the recording format, so that figure is a film
measurement. The catalogue mixes film and digital in one layer — in a sampled
area roughly a fifth of frames are `Digital - Colour` — and a digital frame has
no negative. `fly_footprint()` therefore sizes each row from its `media` value
and records the outcome in `footprint_basis`. Frames whose format it cannot
resolve get an empty geometry rather than a plausible rectangle, and every
downstream function reports how many it excluded:

```r
table(footprints$footprint_basis)
sized <- footprints[footprints$footprint_basis != "unknown_format", ]
```

Digital frames resolve to `"unknown_format"` because the sensor width they need
is not in the centroid metadata. Supply it yourself if you know the camera:
`fly_footprint(photos, format_size = c("Digital - Colour" = 3.54))`.

The photos used throughout this vignette are 1968 film, so every footprint below
is sized from the 9-inch negative.

By default footprints are sized from the reported scale, which assumes flat
ground at whatever elevation that scale was computed for. Supplying a DEM
removes that assumption — worth about 14% of footprint area here, and shown at
the end of this vignette. Every coverage and overlap number below inherits
whichever basis you choose.
Figure \@ref(fig:fig-footprint) shows the estimated footprints for all 20
photos. Notice that some centroids fall outside the AOI while their footprints
still overlap it — `fly_filter()` with `method = "footprint"` catches these
edge cases that a simple centroid-in-polygon filter would miss.

```{r fig-footprint, fig.cap = "Estimated photo footprints (blue rectangles) and centroids (red dots) overlaid on the Upper Bulkley River floodplain AOI."}
footprints <- fly_footprint(centroids)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
plot(st_geometry(footprints), border = "steelblue", add = TRUE)
plot(st_geometry(centroids), pch = 20, cex = 0.5, col = "red", add = TRUE)
```

```{r filter}
fp_result <- fly_filter(centroids, aoi, method = "footprint")
ct_result <- fly_filter(centroids, aoi, method = "centroid")
knitr::kable(data.frame(
  Method = c("footprint", "centroid"),
  Photos = c(nrow(fp_result), nrow(ct_result)),
  Description = c("Footprint overlaps AOI", "Centroid inside AOI")
), caption = "Comparison of spatial filtering methods.")
```

# Summary statistics

`fly_summary()` reports footprint dimensions and date ranges by scale.

```{r summary}
fly_summary(centroids)
```

# Coverage analysis

`fly_coverage()` computes what percentage of the AOI is covered by photo
footprints, grouped by any column.

```{r coverage}
fly_coverage(centroids, aoi, by = "scale")
```

# Photo selection

`fly_select()` has two modes:

- `mode = "minimal"` — fewest photos to reach target coverage
- `mode = "all"` — every photo whose footprint touches the AOI

## Minimal selection

```{r select-minimal}
selected <- fly_select(centroids, aoi, mode = "minimal", target_coverage = 0.80)
selected[, c("airp_id", "scale", "selection_order", "cumulative_coverage_pct")]
```

Figure \@ref(fig:fig-minimal) shows the greedy minimal selection result.

```{r fig-minimal, fig.cap = "Minimal greedy selection — fewest photos to reach 80% AOI coverage."}
sel_fp <- fly_footprint(selected)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
plot(st_geometry(sel_fp), border = "steelblue", col = adjustcolor("steelblue", 0.15), add = TRUE)
plot(st_geometry(selected), pch = 20, col = "red", add = TRUE)
```

## Ensuring component coverage {#ensure-components}

When the AOI has multiple disconnected polygons (e.g. patchy floodplain
fragments), minimal selection optimizes total area and can leave entire
components uncovered. Use `component_ensure = TRUE` to guarantee at least
one photo per component before running greedy selection:

```{r select-components}
# How many polygon components in our AOI?
n_components <- length(sf::st_cast(st_union(aoi), "POLYGON"))
cat("AOI has", n_components, "polygon components\n")

selected_ec <- fly_select(centroids, aoi, mode = "minimal",
                          target_coverage = 0.80, component_ensure = TRUE)
cat("Without component_ensure:", nrow(selected), "photos\n")
cat("With component_ensure:   ", nrow(selected_ec), "photos\n")
```

Compare Figure \@ref(fig:fig-minimal) with Figure \@ref(fig:fig-components) —
the component-ensured selection covers more of the disconnected floodplain
fragments at the cost of a few extra photos.

```{r fig-components, fig.cap = "Component-ensured selection — every disconnected AOI polygon gets at least one photo before greedy backfill."}
sel_fp_ec <- fly_footprint(selected_ec)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
plot(st_geometry(sel_fp_ec), border = "steelblue",
     col = adjustcolor("steelblue", 0.15), add = TRUE)
plot(st_geometry(selected_ec), pch = 20, col = "red", add = TRUE)
```

## All photos touching AOI

```{r select-all}
all_in_aoi <- fly_select(centroids, aoi, mode = "all")
cat(nrow(all_in_aoi), "photos intersect the AOI\n")
```

# Overlap analysis

`fly_overlap()` reports pairwise overlap between photo footprints.
Run it on same-scale subsets to understand coverage quality.

```{r overlap}
photos_12k <- centroids[centroids$scale == "1:12000", ]
overlap_12k <- fly_overlap(photos_12k)
overlap_12k
```

```{r overlap-summary}
if (nrow(overlap_12k) > 0) {
  cat("1:12000 overlap range:",
      round(min(overlap_12k$pct_of_a), 1), "% -",
      round(max(overlap_12k$pct_of_a), 1), "%\n")
}

photos_31k <- centroids[centroids$scale == "1:31680", ]
overlap_31k <- fly_overlap(photos_31k)
if (nrow(overlap_31k) > 0) {
  cat("1:31680 overlap range:",
      round(min(overlap_31k$pct_of_a), 1), "% -",
      round(max(overlap_31k$pct_of_a), 1), "%\n")
}
```

# Multi-scale workflow: best resolution first

In practice you want the finest-scale photos first, then fill gaps with
coarser scales. Sort scales finest-first by parsing the numeric denominator:

```{r multi-scale}
sf_use_s2(FALSE)

# Sort scales finest-first
scales <- sort(unique(as.numeric(gsub("1:", "", centroids$scale))))
cat("Scales (finest first):", paste0("1:", scales), "\n")

target_coverage <- 0.80
aoi_albers <- st_transform(aoi, 3005) |> st_union() |> st_make_valid()
aoi_area <- as.numeric(st_area(aoi_albers))
selected_all <- NULL
remaining_aoi <- aoi_albers

for (sc_num in scales) {
  sc <- paste0("1:", sc_num)
  photos_sc <- centroids[centroids$scale == sc, ]
  if (nrow(photos_sc) == 0) next

  # Take all photos at this scale that touch the (remaining) AOI
  remaining_sf <- st_sf(geometry = st_geometry(remaining_aoi)) |>
    st_transform(4326) |> st_make_valid()
  sel <- fly_select(photos_sc, remaining_sf, mode = "all")
  if (nrow(sel) == 0) next

  # Update remaining uncovered area
  fp <- fly_footprint(sel) |> st_transform(3005)
  fp_union <- st_union(fp) |> st_make_valid()
  remaining_aoi <- tryCatch(
    st_difference(remaining_aoi, fp_union) |> st_make_valid(),
    error = function(e) remaining_aoi
  )
  remaining_area <- as.numeric(st_area(remaining_aoi))
  covered_pct <- 1 - sum(remaining_area) / aoi_area

  cat(sc, ":", nrow(sel), "photos (cumulative coverage:",
      round(covered_pct * 100, 1), "%)\n")

  sel$priority_scale <- sc
  selected_all <- rbind(selected_all, sel)

  if (covered_pct >= target_coverage) break
}

cat("\nTotal:", nrow(selected_all), "photos\n")
as.data.frame(table(selected_all$priority_scale))
```

Figure \@ref(fig:fig-multi-scale) shows the result — finest-scale photos
(blue) provide the primary coverage, with coarser-scale photos (orange)
filling remaining gaps.

```{r fig-multi-scale, fig.cap = "Multi-scale priority selection — finest-scale photos first (blue), coarser scales backfill gaps (orange)."}
sel_fp <- fly_footprint(selected_all)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
scale_labels <- sort(unique(selected_all$priority_scale))
palette <- c("steelblue", "darkorange", "forestgreen", "firebrick")
cols <- palette[match(selected_all$priority_scale, scale_labels)]
for (j in seq_len(nrow(sel_fp))) {
  plot(st_geometry(sel_fp[j, ]), border = cols[j],
       col = adjustcolor(cols[j], 0.15), add = TRUE)
}
legend("topright", legend = scale_labels,
       fill = adjustcolor(palette[seq_along(scale_labels)], 0.3),
       border = palette[seq_along(scale_labels)], bty = "n")
```

# Thumbnail retrieval and georeferencing

`fly_fetch()` downloads thumbnail images (or flight logs, calibration
reports) from the BC Data Catalogue URLs included in the centroid data.
`fly_georef()` warps each thumbnail to its estimated footprint
polygon, producing georeferenced GeoTIFFs in BC Albers.

```{r fetch-georef}
fetched <- fly_fetch(centroids[1:3, ], type = "thumbnail",
                     dest_dir = tempdir())
georef <- fly_georef(fetched, centroids[1:3, ],
                           dest_dir = tempdir())
georef[, c("airp_id", "dest", "success")]
```

The georeferenced TIFFs inherit whatever basis `fly_footprint()` was given,
plus its nadir-camera assumption — `fly_georef()` takes the same `dem`
argument. They are approximate either way, useful for visual context rather
than survey-grade positioning. Metadata from the original centroid data (date,
scale, focal length) links back via `airp_id`.

# Terrain-adjusted footprints

Everything above sizes each frame from its reported scale. That scale is
referenced to some elevation, and where the ground sits below that elevation the
photo covers more than the scale implies.

The catalogue carries what is needed to do better. `FLYING_HEIGHT` is metres
above **sea level**, not height above ground, so subtracting terrain elevation
gives the height ground coverage actually scales with:

```{r terrain-compare}
dem <- system.file("testdata/dem.tif", package = "fly")

flat    <- fly_footprint(centroids)
terrain <- fly_footprint(centroids, dem = dem)

pct <- 100 * (as.numeric(st_area(st_transform(terrain, 3005))) /
                as.numeric(st_area(st_transform(flat, 3005))) - 1)

dplyr::tibble(
  scale        = centroids$scale,
  elevation_m  = round(centroids$flying_height - terrain$height_agl),
  agl_m        = round(terrain$height_agl),
  area_change  = round(pct, 1)
) |>
  dplyr::group_by(scale) |>
  dplyr::summarise(
    n          = dplyr::n(),
    ground_m   = paste0(min(elevation_m), "-", max(elevation_m)),
    `area +%`  = paste0(round(min(area_change), 1), " to ", round(max(area_change), 1)),
    median     = round(median(area_change), 1)
  ) |>
  knitr::kable(caption = "Footprint area change from terrain-adjusted sizing, by scale.")
```

Every footprint grows, and none shrink. That one-directional bias is the tell
that this is not a slope effect: the reported scale is referenced to an elevation
above the valley floor these photos actually cover, so it understates every
frame. Slope would push in both directions.

```{r fig-terrain, fig.cap = "Flat-terrain footprints (grey) against terrain-adjusted footprints (red) for the 1:31680 frames. The correction is a uniform enlargement per frame, not a change of shape."}
idx <- centroids$scale == "1:31680"
plot(st_geometry(terrain[idx, ]), border = "firebrick")
plot(st_geometry(flat[idx, ]), border = "grey50", add = TRUE)
plot(st_geometry(aoi), col = NA, border = "grey20", add = TRUE)
```

`footprint_terrain` records what happened to each frame, so a fallback is
visible rather than silent:

```{r terrain-basis}
table(terrain$footprint_terrain)
```

The bundled `dem.tif` is a clip of **MRDEM-30**, NRCan's 30 m bare-earth DTM.
For your own area of interest, read it straight off the public COG — no download
or account needed, since `/vsicurl/` fetches only the bytes that intersect:

```r
mrdem <- paste0(
  "/vsicurl/https://canelevation-dem.s3.ca-central-1.amazonaws.com/",
  "mrdem-30/mrdem-30-dtm.tif"
)
fly_coverage(centroids, aoi, by = "scale", dem = mrdem)
```

Buffer generously. Resolution matters less than extent here — 30 m resolves a
2.7 km footprint's mean elevation perfectly well, but a DEM that stops short of
the frame edges will not. Only a frame whose *centroid* has no elevation takes
the `no_dem_coverage` fallback; one whose centroid is covered and whose edges
are not is still corrected, from the mean of the covered part. `dem_coverage`
reports the fraction per frame, and anything below 95% warns:

```{r terrain-coverage}
round(range(terrain$dem_coverage), 3)
```

Reprojecting a DEM leaves NA slivers along its edges, which is why these are
just under 1 rather than exactly 1. Allow at least half the widest footprint —
3.6 km at 1:31680 — and neither case arises.

What a DEM does **not** fix is the nadir assumption. The catalogue carries no
tilt, roll or crab, so footprints stay axis-aligned rectangles. On this area
that per-corner refinement is worth roughly 2%, against the 14% the DEM
addresses.

