---
title: "Airphoto Selection Pipeline"
output:
  bookdown::html_document2:
    number_sections: true
    base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Airphoto Selection Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the full `fly` pipeline using bundled test data
from the Upper Bulkley River floodplain near Houston, BC. The data
includes 1968 airphoto centroids at two scales (1:12,000 and 1:31,680).

```{r load}
library(fly)
library(sf)

centroids <- st_read(system.file("testdata/photo_centroids.gpkg", package = "fly"), quiet = TRUE)
aoi <- st_read(system.file("testdata/aoi.gpkg", package = "fly"), quiet = TRUE)
```

# Footprint estimation

`fly_footprint()` converts point centroids into rectangular polygons
representing estimated ground coverage. The standard 9" x 9" (228 mm) negative
used by BC aerial survey cameras produces a footprint width of
`9 * scale * 0.0254` metres. The 9x9 format is the default (`negative_size = 9`)
and is correct for standard BC catalogue photography. The full BC Air Photo
Database records camera focal length per roll, but this field is not available
in the simplified centroid data — override `negative_size` if working with
non-standard format photography.

Note that footprints assume flat terrain beneath the aircraft. On slopes the
true ground coverage differs — downhill slopes produce a larger actual
footprint, uphill slopes a smaller one. All coverage and overlap numbers
downstream inherit this approximation.
Figure \@ref(fig:fig-footprint) shows the estimated footprints for all 20
photos. Notice that some centroids fall outside the AOI while their footprints
still overlap it — `fly_filter()` with `method = "footprint"` catches these
edge cases that a simple centroid-in-polygon filter would miss.

```{r fig-footprint, fig.cap = "Estimated photo footprints (blue rectangles) and centroids (red dots) overlaid on the Upper Bulkley River floodplain AOI."}
footprints <- fly_footprint(centroids)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
plot(st_geometry(footprints), border = "steelblue", add = TRUE)
plot(st_geometry(centroids), pch = 20, cex = 0.5, col = "red", add = TRUE)
```

```{r filter}
fp_result <- fly_filter(centroids, aoi, method = "footprint")
ct_result <- fly_filter(centroids, aoi, method = "centroid")
knitr::kable(data.frame(
  Method = c("footprint", "centroid"),
  Photos = c(nrow(fp_result), nrow(ct_result)),
  Description = c("Footprint overlaps AOI", "Centroid inside AOI")
), caption = "Comparison of spatial filtering methods.")
```

# Summary statistics

`fly_summary()` reports footprint dimensions and date ranges by scale.

```{r summary}
fly_summary(centroids)
```

# Coverage analysis

`fly_coverage()` computes what percentage of the AOI is covered by photo
footprints, grouped by any column.

```{r coverage}
fly_coverage(centroids, aoi, by = "scale")
```

# Photo selection

`fly_select()` has two modes:

- `mode = "minimal"` — fewest photos to reach target coverage
- `mode = "all"` — every photo whose footprint touches the AOI

## Minimal selection

```{r select-minimal}
selected <- fly_select(centroids, aoi, mode = "minimal", target_coverage = 0.80)
selected[, c("airp_id", "scale", "selection_order", "cumulative_coverage_pct")]
```

Figure \@ref(fig:fig-minimal) shows the greedy minimal selection result.

```{r fig-minimal, fig.cap = "Minimal greedy selection — fewest photos to reach 80% AOI coverage."}
sel_fp <- fly_footprint(selected)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
plot(st_geometry(sel_fp), border = "steelblue", col = adjustcolor("steelblue", 0.15), add = TRUE)
plot(st_geometry(selected), pch = 20, col = "red", add = TRUE)
```

## Ensuring component coverage {#ensure-components}

When the AOI has multiple disconnected polygons (e.g. patchy floodplain
fragments), minimal selection optimizes total area and can leave entire
components uncovered. Use `component_ensure = TRUE` to guarantee at least
one photo per component before running greedy selection:

```{r select-components}
# How many polygon components in our AOI?
n_components <- length(sf::st_cast(st_union(aoi), "POLYGON"))
cat("AOI has", n_components, "polygon components\n")

selected_ec <- fly_select(centroids, aoi, mode = "minimal",
                          target_coverage = 0.80, component_ensure = TRUE)
cat("Without component_ensure:", nrow(selected), "photos\n")
cat("With component_ensure:   ", nrow(selected_ec), "photos\n")
```

Compare Figure \@ref(fig:fig-minimal) with Figure \@ref(fig:fig-components) —
the component-ensured selection covers more of the disconnected floodplain
fragments at the cost of a few extra photos.

```{r fig-components, fig.cap = "Component-ensured selection — every disconnected AOI polygon gets at least one photo before greedy backfill."}
sel_fp_ec <- fly_footprint(selected_ec)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
plot(st_geometry(sel_fp_ec), border = "steelblue",
     col = adjustcolor("steelblue", 0.15), add = TRUE)
plot(st_geometry(selected_ec), pch = 20, col = "red", add = TRUE)
```

## All photos touching AOI

```{r select-all}
all_in_aoi <- fly_select(centroids, aoi, mode = "all")
cat(nrow(all_in_aoi), "photos intersect the AOI\n")
```

# Overlap analysis

`fly_overlap()` reports pairwise overlap between photo footprints.
Run it on same-scale subsets to understand coverage quality.

```{r overlap}
photos_12k <- centroids[centroids$scale == "1:12000", ]
overlap_12k <- fly_overlap(photos_12k)
overlap_12k
```

```{r overlap-summary}
if (nrow(overlap_12k) > 0) {
  cat("1:12000 overlap range:",
      round(min(overlap_12k$pct_of_a), 1), "% -",
      round(max(overlap_12k$pct_of_a), 1), "%\n")
}

photos_31k <- centroids[centroids$scale == "1:31680", ]
overlap_31k <- fly_overlap(photos_31k)
if (nrow(overlap_31k) > 0) {
  cat("1:31680 overlap range:",
      round(min(overlap_31k$pct_of_a), 1), "% -",
      round(max(overlap_31k$pct_of_a), 1), "%\n")
}
```

# Multi-scale workflow: best resolution first

In practice you want the finest-scale photos first, then fill gaps with
coarser scales. Sort scales finest-first by parsing the numeric denominator:

```{r multi-scale}
sf_use_s2(FALSE)

# Sort scales finest-first
scales <- sort(unique(as.numeric(gsub("1:", "", centroids$scale))))
cat("Scales (finest first):", paste0("1:", scales), "\n")

target_coverage <- 0.80
aoi_albers <- st_transform(aoi, 3005) |> st_union() |> st_make_valid()
aoi_area <- as.numeric(st_area(aoi_albers))
selected_all <- NULL
remaining_aoi <- aoi_albers

for (sc_num in scales) {
  sc <- paste0("1:", sc_num)
  photos_sc <- centroids[centroids$scale == sc, ]
  if (nrow(photos_sc) == 0) next

  # Take all photos at this scale that touch the (remaining) AOI
  remaining_sf <- st_sf(geometry = st_geometry(remaining_aoi)) |>
    st_transform(4326) |> st_make_valid()
  sel <- fly_select(photos_sc, remaining_sf, mode = "all")
  if (nrow(sel) == 0) next

  # Update remaining uncovered area
  fp <- fly_footprint(sel) |> st_transform(3005)
  fp_union <- st_union(fp) |> st_make_valid()
  remaining_aoi <- tryCatch(
    st_difference(remaining_aoi, fp_union) |> st_make_valid(),
    error = function(e) remaining_aoi
  )
  remaining_area <- as.numeric(st_area(remaining_aoi))
  covered_pct <- 1 - sum(remaining_area) / aoi_area

  cat(sc, ":", nrow(sel), "photos (cumulative coverage:",
      round(covered_pct * 100, 1), "%)\n")

  sel$priority_scale <- sc
  selected_all <- rbind(selected_all, sel)

  if (covered_pct >= target_coverage) break
}

cat("\nTotal:", nrow(selected_all), "photos\n")
as.data.frame(table(selected_all$priority_scale))
```

Figure \@ref(fig:fig-multi-scale) shows the result — finest-scale photos
(blue) provide the primary coverage, with coarser-scale photos (orange)
filling remaining gaps.

```{r fig-multi-scale, fig.cap = "Multi-scale priority selection — finest-scale photos first (blue), coarser scales backfill gaps (orange)."}
sel_fp <- fly_footprint(selected_all)
plot(st_geometry(aoi), col = "lightyellow", border = "grey40")
scale_labels <- sort(unique(selected_all$priority_scale))
palette <- c("steelblue", "darkorange", "forestgreen", "firebrick")
cols <- palette[match(selected_all$priority_scale, scale_labels)]
for (j in seq_len(nrow(sel_fp))) {
  plot(st_geometry(sel_fp[j, ]), border = cols[j],
       col = adjustcolor(cols[j], 0.15), add = TRUE)
}
legend("topright", legend = scale_labels,
       fill = adjustcolor(palette[seq_along(scale_labels)], 0.3),
       border = palette[seq_along(scale_labels)], bty = "n")
```

# Thumbnail retrieval and georeferencing

`fly_fetch()` downloads thumbnail images (or flight logs, calibration
reports) from the BC Data Catalogue URLs included in the centroid data.
`fly_georef()` warps each thumbnail to its estimated footprint
polygon, producing georeferenced GeoTIFFs in BC Albers.

```{r fetch-georef}
fetched <- fly_fetch(centroids[1:3, ], type = "thumbnail",
                     dest_dir = tempdir())
georef <- fly_georef(fetched, centroids[1:3, ],
                           dest_dir = tempdir())
georef[, c("airp_id", "dest", "success")]
```

The georeferenced TIFFs inherit the flat-terrain and nadir-camera
assumptions from `fly_footprint()` — they are approximate, useful for
visual context rather than survey-grade positioning. Metadata from the
original centroid data (date, scale, focal length) links back via
`airp_id`.
